---
title: "My First Rmarkdown"
author: "Christian Supsup"
date: "01 June 2021"
output: html_document
---

# Markdown syntax (Header 1)

## Markdonw syntax (Header 2)

**Bold font**

*Italic font*

## Rmarkdown setup: install required packages
```{r setup}
#install.packages("rmarkdown")
#install.packages(c("rmarkdown", "knitr"))

## Mac requires MacTeX: https://www.tug.org/mactex/?utm_source=chatgpt.com
## Windows requires TeX Live: https://www.tug.org/texlive/?utm_source=chatgpt.com
```

## Code chunks

Some code chunk instructions:

**echo** (echo = TRUE) - is the code displayed alongside the results?

**warning** (warning = TRUE) - are warning messages displayed?

**fig.width, fig.height** (fig.width = 7) - what width/height (in inches) are the plots?

**fig.aling** (fig.align = "left") - left, right, and center alignment


```{r chunks, echo = TRUE, warning = TRUE}
## Run/render rmarkdown
#rmarkdown::render("act06_rmarkdown.rmd")
```

## Inserting figures
```{r figs, fig.width = 4, fig.height = 3}
A <- c("a", "a", "b", "b")
B <- c(5, 10, 15, 20)
df <- data.frame(A, B)
print(df)
boxplot(B~A,data=df)
```

## Inserting tables
```{r tables}
## Using R standard print
df

## Using kable() function
library(knitr)
kable(df, digits = 2)

## Using pander
library(pander)
plant <- c("a", "b", "c")
temperature <- c(20, 20, 20)
growth <- c(0.65, 0.95, 0.15)
df <- data.frame(plant, temperature, growth)
emphasize.italics.cols(3)   ## Make the 3rd column italics
pander(df)           ## Create the table
```

Manually creating a table

| Plant | Temp. | Growth |
|:------|:-----:|-------:|
| A     | 20    | 0.65   |
| B     | 20    | 0.95   |
| C     | 20    | 0.15   |








