---
title: "Activity 07 - Clustering (K-means)"
author: "Jerome Poblete"
date: "2026-06-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# A. K-means (single K)
## 1. Read in the morphology data as a dataframe.
```{r}
df <- read.csv("morphology_data.csv")
```

## 2. Remove the sex column.
```{r}
df.no_sex <- subset(df, select = -Sex)
```

## 3. Ensure that the dataframe has no NAs.
```{r}
df.no_sex <- df.no_sex[complete.cases(df.no_sex), ]
```

```{r}
morph <- df.no_sex[2:13]
```

## 4. Prior to performing K-means, standardize all values using the “scale” function.
```{r}
morph.scaled <- scale(morph)
```

## 5. Perform K-means with 3 cluster centers using 25 initial random centroids.
```{r}
# Ensure our results are reproduceable.
set.seed(1234567890)
morph.km <- kmeans(morph.scaled, centers=3, nstart=25)
```

## 6. Show the cluster assignment and a summary of results by simply printing them to the console.
```{r}
head(morph.km$cluster)
```

```{r}
morph.km
```

## 7. Create a scatter plot with points colored by the assigned cluster.
```{r}
# Use the first 2 variables: x=SVL, y=TD. Note that since morph.scaled is a
# matrix, we need to subset it as such to access the column-wise arrays.
plot(morph.scaled[, 1:2],
     main="K-means Clustering",
     xlab="SVL", ylab="TD",
     col=morph.km$cluster)

# Add centroids to the previous scatter plot.
points(morph.km$centers[, 1:2], col=1:3, pch=8, lwd=2)
```

# B. K-means (loop)
## 1. Now, create a loop function to perform K-means with cluster centers from 2 to 10.
```{r}
centers <- 1:10
ss <- vector(mode="numeric", length=length(centers))
for (i in centers) {
  km <- kmeans(morph.scaled, centers=i, nstart=1)
  ss[i] <- km$tot.withinss
}
```

## 2. From the output of your loop function, create an elbow curve plot to determine the optimal K.
```{r}
plot(centers, ss,
     main="Elbow Curve",
     type="b", xlab="K values",
     ylab="Total Within SS")
```

Based on the above scatter plot, my (subjective) observation is that the elbow is 3.

## 3. Recreate steps 6–8 above using the optimal K determined by the elbow curve.
```{r}
morph.km2 <- kmeans(morph.scaled, centers=3, nstart=1)
plot(morph.scaled[, 1:2],
     main="K-means Clustering from Elbow",
     xlab="SVL", ylab="TD",
     col=morph.km2$cluster)

# Plot centroids to the previous scatter plot.
points(morph.km2$centers[, 1:2], col=1:3, pch=8, lwd=2)

```