---
title: "Acitivy 03: Summary statistics, frequency distribution, and basic graphical representation"
author: "Jerome Poblete"
date: "28 May 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE)

```

# A. Read and perform descriptive statistics

## 1. Read in the CSV file

We use the base R function `read.csv()`, which returns a `data.frame`.
The data is cleaned prior to performing any tasks.
We do **not** clean the data in Excel but rather do it all in R via its subset
operations.

```{r df-init}
# NOTE: Ensure the CSV file is in the parent directory of this HTML!
df <- read.csv("../herpetozoa-033-095-s001.csv")
df <- df[complete.cases(df), ]
```

Recall than in R, subsets operate on a row, column basis.
Think of this like y, x rather than x, y.
Also recall than these 'indexes' can be logical data frames themselves.
So the line `df[complete.cases(df), ]` selects all rows with non-empty data
and selects all their columns as we assume all column headers are valid.

## 2. Show the data structure

Printing builtin data structures in simple enough. Their type name can be
retrieved via `class()` and a simple representation via `str()`. If there is
a non-empty return value, R implicitly prints that. There is no need to
explicitly call `print()`.

```{r df-show}
class(df)
str(df)
```

## 3. Subset the elevation data

Like before we use the R subset operations to filter from the given data frame.
We are selecting all rows of the `Elev` column specifically.

```{r df-subset-elev}
elev.data <- df[, "Elev"]
elev.data
```

We can also provide the following summary statistics:
  
```{r elev-stats}
mean(elev.data)
sd(elev.data)
range(elev.data)
```

## 4. Subset the species data

# B. Graphical representation

```{r graph-init, echo=FALSE}
species.min = "Babu"
species.max = "Phlo"
```

## 1. Scatterplot

We create a scatter plot of total species count, elevation, and habitat types
using the base R `plot()`function. We also colorize it for better visualization.

```{r fig.cap="Elevation and total species count."}
htypes.colors_map <- c(ASG="green", ESG="blue", CVT="red", OGF="orange")
htypes.colors <- sapply(df[, "Htypes"], function(k) {htypes.colors_map[k]})
plot(x    = df[, "Elev"],
     y    = df[, "SPPTotal"],
     xlab = "Elevation",
     ylab = "Species total count",
     col  = htypes.colors,
     bg   = htypes.colors,
     pch  = 17)
```

## 2. Bar plot

We create a bar plot of habitat type and species count using the base R
`barplot()` function.

```{r, fig.cap="Habitat type and total species count."}
d <- df[, c("SPPTotal", "Elev", "Htypes")]
barplot(
  height = sapply(split(d, d$Htypes), function(v) sum(v$SPPTotal)),
  xlab   = "Habitat type",
  ylab   = "Total species count"
)
```

## 3. Histogram

We create a histogram of the frequency of each species total count using the
base R `hist()` function.

```{r, fig.cap="Species total count frequency."}
hist(df[, "SPPTotal"], xlab="Species count total", main="Species count total frequency")
```

# C. Interpretation
> Based on the summary statistics and preliminary graphs you just created,
> how do you describe the relationship between the following:
>
> 1. total species observed
> 2. habitat types
> 3. elevation

The total count of species observed seems to have a very loose inverse
relationship with elevation.
That is, in the Elevation-Species total count scatter plot, increasing elevation
leads to decrease in species count.
Species also seem to be most abundant at ASG and least abundant at CVT.
This may explain by lower counts of species are more common than higher ones as
seen in the histogram.