---
title: "Activity 04: Tidyverse"
author: "Jerome Poblete"
date: "25 May 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
```

# A. Read and filter out
## 1. Install tidyverse and load

First load the `tidyverse` library.

```{r ht-init, message=FALSE}
# install.packages("tidyverse")
library(tidyverse)
```


## 2. Read in the CSV file as a data frame

Load the habitat types data set using the base R `read.csv()` function.

```{r ht-read}
# NOTE: Ensure this file is found in the parent directory of this HTML.
ht <- read.csv("../habitat_types.csv")
```

## 3. Show the data using the tidyverse equivalent of base R `str`

Here we are introduced to the R pipe operator `|>`.
There is the more archaic `%>%` but the modern version `|>` is more succinct.
It allows us to pass the left hand side as the first argument to the function
call on the right hand side.

So the following code (and output):

```{r}
ht |> glimpse()
```

Is equivalent to:

```{r, eval=FALSE}
glimpse(ht)
```

The pipe operator is arguably cleaner and aids in readability when many
function calls are chained together but constantly operate on the same input.
If you're curious, read on 'Tidy Evaluation'.

## 4. Subset the data with some elect columns

We are also introduced to the Tidyverse concept of 'quosures'. These are
unquoted strings which can implicitly refer to the main data set's fields.
That is, instead of constantly writing `ht$Htypes`, Tidyverse functions can
simply receive the identifier `Htypes` or the string `"Htypes"` and work out
that you want the field of `ht` as it was passed as the first argument.

So the following code:

```{r}
ht2 <- ht |> select(Transect, Htypes, SmallTrees, FloweringTrees)
```

Is equivalent to both of the following:

```{r, eval=FALSE}
ht2 <- select(ht, Transect, Htypes, SmallTrees, FloweringTrees)
ht3 <- select(ht, "Transect", "Htypes", "SmallTrees", "FloweringTrees")
```

## 5. Filter by habitat type and sort by small trees

The dplyr `filter()` function also allows you to use 'quosures'. And just
like in base R, the subset can be a logical expression so that you extract
only the entries that meet the specified condition.

The dplyr `arrange()` function works similarly. The argument is a lazily
evaluated expression. See the dplyr function `?desc`.

So the following code:

```{r}
cvt <- ht2 |> filter(Htypes == "CVT") |> arrange(desc(SmallTrees))
```

Is equivalent to:

```{r, eval=FALSE}
tmp <- filter(ht2, ht2$Htypes == "CVT")
cvt <- arrange(tmp, desc(tmp$SmallTrees))
```

Hopefully you can now see why the pipe operator is useful. Without it,
we had to save the result of `filter()` in a variable in order to
pass it to `desc()` and `arrange()`! Otherwise you'd need to copy the call
to `filter()` multiple times.

We can then find out which transects have the highest counts of small trees.

```{r}
cvt
```

In the above table, we can see that the highest count of small trees is ten (10)
from transects IT2P7 and IT3P0.

## 6. Calculate the mean number of flowering trees per habitat

A new `data.frame` is created using the dplyr `summarize()` function.
All named arguments (aside from `.data`, .by` and `.groups`) represent the
row name and column data to be used.

```{r}
ftrees <- ht2 |> group_by(Htypes) |> summarize(mean_ft=mean(FloweringTrees))
ftrees
```

The habitat types that do not differ significantly in the mean number of
flowering trees are ASG and ESG, at 5.26 and 5.38 flowering trees,
respectively. CVT on the other hand has an average of 8.86 flowering trees
 making it more than 3.00 higher than the other 2.

## 7. Add a new column for the difference between herb and ground cover

We use the dplyr `mutate()` function to add a new column to the input data.
The original data is modified, which is why we created `ht2` to avoid messing
with `ht`.

```{r}
sel_df_mod <- ht2 |> mutate(HerbGroundDiff = ht$HerbsCover - ht$GroundCover)
```

The following transects have no difference:

```{r}
sel_df_mod |> filter(HerbGroundDiff == 0)
```

## 8. Convert `sel_df_mod` to a wider output

The dplyr `pivot_wider()` function allows you to 'expand' the given input.
Providing the named argument `names_from` will provide the suffix for the
column titles for the values in `values_from`.

Each column titel is a combination of a value in `values_from` and `names_from`
Many values here will be `NA` as each transect has only 1 habitat types. So
the combinations with the non-existent habitat type, e.g. IT1P0 with CVT will
be `NA`.

```{r}
sel_df_mod |> pivot_wider(
  names_from  = Htypes,
  values_from = c(SmallTrees, FloweringTrees, HerbGroundDiff, SmallTrees))
```

# B. Reflection
> How does using tidyverse compare to Activity 03, which primarily used 
> Base R? Which functions do you find more intuitive, and which ones are
> less so?

I felt that the data frame manipulation functions were a lot easier to
follow especially since we now know about the pipe operator. `filter()`,
`arrange()`, `group_by()` and `summarize()` are very helpful since in
Activity 03 I was confused on how to achieve the same results in plain R.

However I'm still trying to wrap my head around `pivot_longer()` and
`pivot_wider()` because they do so much, kind of like the plot functions
we had to to learn how to use back in Activity 03. I think with a bit more
practice these will become more intuiive.