---
title: "Activity 14 - Gradient Boosting"
author: "Faith Lasquite & Jerome Poblete"
date: "2026-07-22"
output: html_document
---

```{r setup, include=FALSE}
rm(list = ls())
knitr::opts_chunk$set(echo = TRUE)
```

## 1) Define gradient boosting & explain how it differs from random forest.

In order to define gradient boosting, it helps to define some terms.


### 1.1) Decision Tree

All of the machine learning we will do will create so-called *Decision Trees*.
These are flowchart-like structures that help us classify a particular point in
a data set.
[(IBM Research, 2026)](https://www.ibm.com/think/topics/decision-trees)

It consists of a series of nodes, which represents a characteristic with a  cutoff point.
We start with the first node, the *Root Node*, which represents the first characteristic we're after.
If the data point's characteristic is  lower or higher/equal-to this characteristic, we proceed to the next appropriate node (i.e. we "decide" to go "left" or "right").
This process continues until we hit the *Leaf Node*, which is the final characterization of the data point.

> **NOTE:** This only works if all characteristics are numeric!


### 1.2) Ensemble Learning
This is technique for machine learning which aggregates (i.e. combines)  two-or-more (2/+) learners, such as regression models and/or neural networks, to produce better predictions.

Why would we need this? Because an individual learner may not be accurate, it could over fit or under fit to the training data.
By combining multiple of them, we hope to increase the overall accuracy.

For our purposes, our *Learners* will be *Decision Trees*. This applies because *Decision Trees* can also over or under fit to the training data, so they can also benefit from this.


#### 1.2.1) Fragmentation

*Ensemble Learning* is useful for *Decision Trees* because large trees can become *fragmented*, where there are many *Leaf Node* classifications with few data points in them.
In the worst case, every single data point has its own classification.
That wouldn't be very useful, now would it?


#### 1.2.2) Overfitting
Fragmentation can lead to the model fitting too closely, or even exactly, to its training data. [(IBM Research, 2026)](https://www.ibm.com/think/topics/overfitting)
This is known as *overfitting*.
It makes the model less useful for actually predicting things, because once it gets data points outside of the training data set, said predictions will be much less accurate.

*Overfitting* can occur when the *Machine Learning model* trains for too long on some sample data, or is too complex.
In these cases it "learns" the noise (i.e. irrelevant information) and makes it hard to generalize for new data.

Some tell-tale signs of *overfitting* include:

- low error rates
- high variance

One (1) way to check for *overfitting* is to use a subset of the training data set as a "test set".
When the training data set has a low error rate but the test data set has a high error rate, this probably means *overfitting* occurred.


### 1.3) Random Forest

This is a form of *Bootstrap Aggregating*, also known as *Bagging*,
which is an example of an *Ensemble Learning* method.


#### 1.3.1) Bagging

This is an *Ensemble Learning* method that aims to **reduce variance** within a "noisy" data set.
[(IBM Research, 2026)](https://www.ibm.com/think/topics/bagging)

Here, a sample of the training data set is taken with replacement- i.e. individual data points can be selected more than once.
This continues until we have several samples, after which we train several independent weak models.
We then get the *average* or *majority* of these weak models to yield more accurate estimates.

There are three (3) basic steps to *Bagging*:

##### 1.3.1.1) Bootstrapping
Select data points at random, and with replacement.
This ensures we have diverse samples.
Note that, since this samples at random *with replacement*, then we are able to select the same data point multiple times in a sample.


##### 1.3.1.2) Parallel training
Bootstrapped samples are independently trained, and at the same time.

##### 1.3.2.3) Aggregation
Get the average or majority of the predictions, depending on the task at hand.

For *Regression*, get the average of all outputs predicted by the individual trees.
This process is known as *Soft Voting*.

For *Classification*, get the class with the highest majority of votes.
This process is known as *Hard Voting* or *Majority Voting*.


#### 1.3.2) Random Forest

Here, multiple independent decision trees are created and merged.
The trees are independent in that they use unique subsets of the training data, and none of them will impact any other in any way.
[(IBM Research, 2026)](https://www.ibm.com/think/topics/random-forest)

Specifically, it is an extension of *Bagging* which creates an uncorrelated "forest" of *Decision Trees*.
It requires *Feature Randomness* (a.k.a. *Feature Bagging*) to generate random subsets of features in order to ensure low correlation among the *Decision Trees*.

*Decision Trees* consider all possible feature splits but *Random Forest* only selects a subset of those features.

Three (3) main hyperparameters need to be set before training:

1) Node size
2) Number of trees
3) Number of features sampled


### 1.4) Gradient Boosting

Like *Random Forest*, this is also an *Ensemble Learning* method that combines multiple *Decision Trees* into one (1) model.
However, it uses the *Boosting* method rather than *Bagging*.


#### 1.4.1) Boosting

Recall that in *Bagging*, weak learners are independent but trained in parallel (i.e. at the same time).
In *Boosting*, however, they are trained sequentially- i.e. one after the other.

Here, a series of models in constructed.
Each each new model iteration, the weights of the misclassified data from the previous are increased.
The reason for this is that it helps the model identify which parameters it needs to focus on in order to improve its performance.

Also, *Bagging* learners are weak, exhibiting high variance with low bias (i.e. low error rates).
In *Boosting*, however, is often used when low variance and high bias is observed.


#### 1.4.2) Gradient Boosting

This is a form of *Boosting* where the residual errors of the previous model are corrected by the next model.
[(IBM Research, 2026)](https://www.ibm.com/think/topics/bagging)

It combines multiple weak prediction models into a single one, i.e. an "ensemble".
These weak models are often *Decision Trees*.
These *Decision Trees* are trained sequentially to minimize errors in order to improve accuracy.
The combination of multiple decision tree regressors or classifiers helps capture complex relationships between features.

The main benefit is to iteratively minimize the *Loss Function* to improve predictive accuracy.
One (1) example of a *Loss Function* is *Mean Squared Error*, given by the formula:

```
MSE = SSE / n

where

SSE = 0
for i in 1:n {
    SSE += (y_i - p_i)^2
}

where

y_i = actual value for observation i
p_i = predicted value for observation i
n   = number of observations
```

There are seven (7) steps to *Gradient Boosting*.

##### 1.4.2.1) Initialization 

##### 1.4.2.2) Calculating residuals

Subtract predicted value from the actual value, for each training sample.
This identifies the areas where the model predictions need improvement.

##### 1.4.2.3) Refining with regularization
##### 1.4.2.4) Training next model
##### 1.4.2.5) Ensemble updates
##### 1.4.2.6) Repetition
##### 1.4.2.7) Stopping criteria

### 1.5) Summary

#### 1.5.1) Bagging vs. Boosting


|                 | Bagging                   | Boosting              |
|-----------------|---------------------------|-----------------------|
| Model sequence  | Parallel, independent     | Sequential, dependent |
| Variance        | High                      | Low                   |
| Bias            | Low                       | High                  |
| Computations    | Expensive                 | Efficient             |
| Overfitting     | Avoids                    | Prone                 | 
| Bias            | Prone                     | Reduces               |

#### 1.5.2) Random Forest vs. Gradient Boosting

|                 | Random Forest             | Gradient Boosting     |
|-----------------|---------------------------|-----------------------|
| Ensemble method | Bagging                   | Boosting              |

---

## 2) Then, similar to random forest, enumerate the implementation types of gradient boosting.
Also provide a brief description of each.

These are popular implementations of *Gradient Boosting* available for both Python and R.

### 2.1) Extreme Gradient Boosting (XGBoost)

[XGBoost](https://xgboost.readthedocs.io/) is an open-source library for *Gradient Boosting* which uses *Gradient Descent*.
It is known for its speed, efficiency, and portability.
Is is especially known for the ability to scale with large data sets.

It is available for [Python](https://xgboost.readthedocs.io/en/stable/install.html#python), and [R](https://xgboost.readthedocs.io/en/stable/install.html#r).

### 2.2) Light Gradient-Boosting Machine (LightGBM)

[LightGBM](https://lightgbm.readthedocs.io/en/latest/index.html) is another open-source library for *Gradient Boosting* which uses *Gradient-based One-Side Sampling* (GOSS)
[(IBM Research, 2026)](https://www.ibm.com/think/topics/gradient-boosting#179371087).

It is best known for its suitability to large-scale data processing as GOSS reduces the computational resources needed by finding split points needed to filter out the data instances.
    

## 3) Pick one (1) type of gradient boosting implementation.
Using AI, study the steps for implementing it.

```{r lightgbm-installation}
################################################################################
## Installation
##
## https://lightgbm.readthedocs.io/en/latest/R/index.html#installing-the-cran-package
################################################################################
## install.packages("lightgbm", repos = "https://cran.r-project.org")
library(lightgbm)
```

For training, we will be using the `agaricus.train` data set provided by the `lightgbm` package.
Before we continue, it helps to understand what exactly the data set contains.

It has two (2) parts: the data itself, the labels.

1) `agaricus.train$data`
A sparse matrix.
Columns are various features of the *Agaricus sp.* mushroom.
Rows are the integers 0 or 1 indicating the presence of said feature.

For example, a particular mushroom may have `cap-shape=bell` set to `0` to indicate its cap shape is not a bell.
All columns of the name `cap-shape=*`, except for one (1), should be set to `0`.
Only a single `cap-shape=*` should be set to `1`.
Think about it, how could a mushroom be simultanously bell-shaped and flat, for example?

Because of the data's binary nature, we should use binary methods.

2) `agaricus.train$label`
A vector of 0's or 1's indicating whether the indexed mushroom is edible or not.


```{r lightgbm-training, warning = FALSE}
################################################################################
## 1) Basic Training
##
## https://github.com/lightgbm-org/LightGBM/blob/main/R-package/demo/basic_walkthrough.R
################################################################################
data(agaricus.train, package = "lightgbm")
data(agaricus.test,  package = "lightgbm")

dtrain <- lgb.Dataset(agaricus.train$data, label = agaricus.train$label)
dtest  <- lgb.Dataset.create.valid(dtrain,
    data  = agaricus.test$data,
    label = agaricus.test$label)

train_params <- list(num_leaves = 4L,
    learning_rate    = 1.0,
    objective        = "binary", # default: "regression"
    min_data_in_leaf = 10L)


## May give a warning from `.get_default_num_threads()` when package
## `RhcpBLASctl` is not installed. It's optional, so nothing should break.
bst1 <- lightgbm(data = dtrain,
    params  = train_params,
    nrounds = 2L,
    ## -1 = no info/debug/warning messages
    ##  0 = no info/debug messaages, still prints warnings
    ##  1 = print evaluation metric
    ##  2 = print evaluation metric and information about tree
    verbose = -1L)

################################################################################
## 2) Basic Prediction
################################################################################
pred1 <- stats::predict(bst1, agaricus.test$data)
err1  <- mean(as.integer(pred1 > 0.5) != agaricus.test$label)
```

Based on the above, the test data's predicted error rate is `r err1`.

```{r lightgbm-validation}
################################################################################
## 3) Validation
################################################################################
valids <- list(train = dtrain, test = dtest)

## `eval` is optional if you want to change the evaluation metrics.
bst2 <- lgb.train(train_params, dtrain,
    nrounds = 2L,
    valids  = valids,
    eval    = c("binary_error", "binary_logloss"), # default: NULL (regression)
    verbose = -1L
)

label <- lightgbm::get_field(dtest, "label")
pred2 <- stats::predict(bst2, agaricus.test$data)
err2  <- as.numeric(sum(as.integer(pred2 > 0.5) != label)) / length(label)
```

Based on the above, the validated error rate is `r err2`.

```{r lightgbm-boosting-prediction}
################################################################################
## LightGBM: Boosting from an initial base prediction
################################################################################
valids <- list(eval = dtrain, test = dtest)

## Start from an initial base prediction
bst3 <- lgb.train(train_params, dtrain, verbose = -1L, valids = valids)

## We need the margin value instead of the transformed prediction as in
## `set_init_score`.
ptrain <- stats::predict(bst3, agaricus.train$data, type = "raw")
ptest  <- stats::predict(bst3, agaricus.test$data,  type = "raw")

lightgbm::set_field(dtrain, "init_score", ptrain)
lightgbm::set_field(dtest,  "init_score", ptest)

## Result of boost from initial prediction
bst4 <- lgb.train(train_params,
    data    = dtrain,
    nrounds = 5L,
    valids  = valids,
    verbose = -1L)

pred_vec <- stats::predict(bst4, agaricus.test$data)
err3     <- mean(as.numeric(pred_vec > 0.5) != agaricus.test$label)
```

Based on the above, the error rate from boosting with an initial base prediction is `r err3`.


## 4) Lastly, implement the steps manually in R without any packages.

### 4.1) Clean up the data
```{r manual-gradient-boosting-init}
data(penguins)

## Remove incomplete data rows.
clean_penguin_df <- na.omit(penguins)

## Randomize all rows, and split them for testing and training.
set.seed(69420)
random_penguin_df <- clean_penguin_df[clean_penguin_df |> nrow() |> sample(), ]

## Number of observations in the original, full data.
n <- nrow(random_penguin_df)

## Parition data set into the first 70% (training) and the last 30% (test).
ofs           <- as.integer(n * 0.7)
train_penguin <- random_penguin_df[1:ofs, ]

## Order of operations! `ofs + 1:n` is parsed as `ofs + (1:n)`.
test_penguin  <- random_penguin_df[(ofs + 1):n, ]
```

### 4.2) Assign the variables
```{r manual-gradient-boosting-assign}
make_df <- function(x_df, y_vec) {
    return(data.frame(Body_Mass = y_vec,
        Bill_Length    = x_df$bill_len,
        Bill_Depth     = x_df$bill_dep,
        Flipper_Length = x_df$flipper_len))
}

## Our training predictors.
x_df <- train_penguin[, c("bill_len", "bill_dep", "flipper_len")]

## Our training response variable.
y_vec <- train_penguin$body_mass

## Number of observations in the training data.
n <- nrow(train_penguin)

## Out training data frame for gradient boosting.
mgb_df <- make_df(x_df, y_vec)

## Our initial prediction is just the mean repeated n times.
pred0         <- mean(y_vec)
pred_curr_vec <- rep(pred0, n)
```


### 4.3) Gradient Boosting loop
First, we need to construct the trees.
```{r manual-gradient-boosting-loop-create-trees}
## 100 is often used, but you can decrease this to reduce computational
## complexity at the cost of accuracy.
n_trees <- 50

## 10% is often used.
learning_rate <- 0.1

## List of trees.
models_vec <- list()
for (i in 1:n_trees) {
    ##Residual for each row
    mgb_df$Residuals <- mgb_df$Body_Mass - pred_curr_vec
    
    ## Note that this assumes the formula variables are also column names.
    int_tree <- lm(Residuals ~ Bill_Length + Bill_Depth + Flipper_Length,
        data = mgb_df)
    
    ## Delegates to `stats::predict.lm()`.
    pred_next_vec <- predict(int_tree, newdata = mgb_df)
    
    ## Update the predictions, scaled by the learning rate.
    pred_curr_vec   <- pred_curr_vec + (learning_rate * pred_next_vec)
    models_vec[[i]] <- int_tree
}
```


Now we can combine the trees, sequentially, to predict the data.
```{r manual-gradient-boosting-loop-predict-data}
mgb_predict <- function(new_data_df, models_vec, pred0, learning_rate) {
    n        <- nrow(new_data_df)
    pred_vec <- rep(pred0, n)
    for (model in models_vec) {
        ## Delegates to `stats::predict.lm()`.
        predN    <- predict(model, newdata = new_data_df)
        pred_vec <- pred_vec + (learning_rate * predN)
    }
    return(pred_vec)
}

final_pred_vec <- mgb_predict(mgb_df, models_vec, pred0, learning_rate)
```

```{r echo = FALSE}
## How many elements from both the head and the tail to dump.
n_end <- 4L
```

We can now compare what the training data predicted to the actual body mass.
Here are the first `r n_end` and last `r n_end` elements for brevity.
```{r echo = FALSE}
gb_dump <- function(pred_vec, actual_df) {
    tmp_df <- data.frame(Predicted = pred_vec, Actual = actual_df)
    n      <- nrow(tmp_df)
    heads  <- 1:n_end
    tails  <- (n - n_end):n
    knitr::kable(tmp_df[union(heads, tails), ], align = 'l')
}

gb_dump(final_pred_vec, y_vec)
```


### 4.4) Training on the new data set
```{r manual-gradient-boosting-loop-train-new}
## Our test predictors.
x2_df  <- test_penguin[, c("bill_len", "bill_dep", "flipper_len")]

## Our test response.
y2_df  <- test_penguin$body_mass

## Number of test observations.
n2     <- nrow(test_penguin)

## Our test data frame for gradient boosting.
mgb2_df <- make_df(x2_df, y2_df)

final_pred2_vec <- mgb_predict(mgb2_df, models_vec, mean(y2_df), learning_rate)
```

Let's compare the first `r n_end` and last `r n_end` predicted test data points to their actual values.
```{r echo = FALSE}
gb_dump(final_pred2_vec, y2_df)
```

Notice how the predicted values for both the training and test data sets are very different from their actual values.
This indicates that the penguin's bill length, bill depth, and flipper length are not accurate markers for tis overall body mass across different species.