---
title:  "Activity 09: Dimension Reduction: Principal Components Analysis (PCA)"
author: "Jerome Poblete"
date:   "2026 June 18"
output: html_document
---
# A. Principal Components

## 2. Read in the frogs' data as a dataframe

```{r read-in}
frog.data <- read.csv("../data/Meneses_et.al_2026_Platymantis.csv")
```

## 3. Subset your data to retain only the "Species" column and morphological features
E.g. SVL, HL, SNL, etc., from columns 10 through the last one (i.e. 10:X).

```{r subset-data}
frog.numeric <- frog.data[, 10:ncol(frog.data)]
```

## 4. Check the class & structure of the subset data. How many morphological features are present in the dataset?

```{r check-class}
library(tidyverse)

class(frog.numeric)
structure(frog.numeric)
frog.morph <- frog.numeric |> select(10:ncol(frog.numeric))
length(frog.morph)
```

There are 16 morphological feature present in the data set.

## 5. Ensure the data frame has no NAs.

```{r ensure-data}
frog.numeric <- frog.numeric[frog.numeric |> complete.cases(), ]
frog.morph   <- frog.morph[frog.morph |> complete.cases(), ]
```

## 6. Scale the data using only the numeric columns, then compute the pairwise correlations. Do not further subset the data.

```{r scale-data}
frog.scaled <- frog.morph  |> scale()
frog.cmat   <- frog.scaled |> cor() |> round(2)
```

## 7. Use the correlation matrix to...

### i. Calculate the following:

1. Eigenvectors

```{r}
frog.eig <- frog.cmat |> eigen()
frog.eig
```

2. Proportion of variance (percentage)

```{r}
pvar <- function(x) {
  props <- x / sum(x)
  data.frame(Proportion = props, Percentage = props * 100)
}

frog.pvar <- pvar(frog.eig$values)
frog.pvar |> round(4)
```

### ii. How many principal components were produced?

```{r}
frog.scores      <- frog.scaled %*% frog.eig$vectors
frog.scores.head <- frog.scores |> head()
frog.scores.head |> round(4)
frog.scores.head |> nrow()
```

6 principals components were produced.

### iii. What is the proportion of variance explained by components 1-4?

```{r}
frog.pvar$Percentage[1:4] |> round(2)
```

Components 1 through 4 explain 33.91%, 16.09%, 11.18%, and 9.34% of the
variance, respectively.

## 8. Create a 1x2 plot showing screeplot & the PCA biplot
You can use either base R or ggplot2.
You can be creative here by coloring the point shapes per species and adding a legend
Consider the high data-ink ratio you learned from the lecture

### i. Scree Plot

```{r}
library(ggplot2)

# NOTE: Using the raw proportions vs. percentages will greatly influence
#       what a good bin width would be!
pcs <- paste0("PC", 1:ncol(frog.morph))

ggplot(frog.pvar, aes(Proportion)) +
  geom_histogram(binwidth = 0.05, fill = "blue")
  
barplot(frog.pvar$Proportion,
        names.arg = pcs,
        ylim = c(0, 1),
        col  = "steelblue",
        main = "Explained Variance (Scree Plot)",
        ylab = "Proportion of Variance")
```

### ii. PCA biplot

```{r}
# What the hell are we gonna use as our alternative for iris$Species
# It must be as long as scores[, 1]
plot(frog.scores[, 1], frog.scores[, 2],
     xlab = "PC1", ylab = "PC2",
     main = "PCA Score Plot (PC1 vs. PC2)",
     pch  = 19,
     col  = factor(frog.data$Species))

legend("topright",
       legend = levels(factor(frog.data$Species)),
       col = 1:3,
       pch = 19)
```

# B. Broken-stick vs. PCAtest

## 1. Determine the variance explained by principal components

### i. Broken-stick method

### ii. PCAtest package

## 2. Create a barplot showing the results of Broken-stick analysis
Also perform a significance test for principal components using PCAtest,
following the instructions at https://arleyc.github.io/PCAtest/.
