---
title: 'Activity 08: Hierarchical Clustering'
author: "Jerome Poblete"
date: "2026-06-11"
output: html_document
---

```{r setup, include=FALSE}
library(knitr)
library(rmarkdown)
knitr::opts_chunk$set(echo = TRUE)
```

# A. Hierarchical clustering
## 1. In RStudio, create an R Markdown file. Name it “Your Surname_hclust.rmd.”
See this file!

## 2. Read in the morphology data as a dataframe.
```{r data-read}
md.raw <- read.csv("../data/morphology_data.csv")
```

## 3. Remove unnecessary columns (e.g., sex, pop).
```{r data-clean}
library(tidyverse)
md.numeric <- select(md.raw, !c("Pop", "Sex"))

# Species data measurements are located at these columns.
md.numeric <- md.numeric[, 1:12]
```

## 4. Ensure that the dataframe has no NAs.
```{r data-clean2}
md.numeric <- md.numeric[complete.cases(md.numeric), ]
```

## 5. Before performing hierarchical clustering, calculate the “Euclidean distance” and use it as your input.
```{r data-euclidean}
md.scaled <- scale(md.numeric)
md.dists  <- dist(md.scaled)
md.names  <- rownames(md.scaled)
```

## 6. Perform hierarchical clustering using Ward’s and Single Linkage methods.
### i. Single Linkage method
```{r data-hierarchical-single}
hc_single <- hclust(md.dists, method="single")
plot(hc_single,
     main="Single Linkage",
     labels=md.names, cex=0.5)
```

### ii. Ward's Method
```{r data-hierarchical-ward}
hc_ward <- hclust(md.dists, method="ward.D2")
plot(hc_ward,
     main="Ward D2",
     labels=md.names, cex=0.5)
```

## 7. Create a for loop function to explore Ks from 2 to 10 for each method using the Silhouette score and display the results with a plot.
```{r data-silhoutte}
explore <- function(data, pretty_name) {
  avg_sil_widths <- vector(mode="numeric", length=9)
  for (i in seq_along(avg_sil_widths)) {
    # Returns a k x 3 matrix with attributes.
    k   <- i + 1
    sil <- silhouette(cutree(data, k), md.dists)
    avg_sil_width <- mean(sil[, 3])
    print(avg_sil_width)
    avg_sil_widths[i] <- avg_sil_width
  }
  plot(2:10, avg_sil_widths, type="b", # Render both points and lines.
       main=paste("Silhouette method (", pretty_name, ")", sep=""),
       xlab="Cluster count (k)", ylab="Average Silhouette width",
       cex=0.5)
}

explore(hc_single, "Single Linkage")
explore(hc_ward, "Ward's Method")
```

## 8. Now, generate a dendrogram for each method, highlighting the optimal K with colored boxes.
```{r data-diana, eval=FALSE, include=FALSE}
# install.packages("cluster)
library(cluster)
diana_fit <- diana(md.scaled)
plot(diana_fit, main="DIANA")
```

### i. Single Linkage
```{r data-dendo-single}
plot(hc_single,
     main="Single Linkage",
     labels=md.names, cex=0.5)

# CMYK :)
rect.hclust(hc_single, k=4, border=c("cyan", "magenta", "yellow", "black"))
```

### ii. Ward's Method
```{r data-dendo-wards}
plot(hc_ward,
     main="Ward D2",
     labels=md.names, cex=0.5)

# CMYK plus one other thing
rect.hclust(hc_ward, k=5, border=c("cyan", "magenta", "yellow", "black", "green"))
```

# B. K-means vs. Hierarchical
## 1. Revisit your K-means outputs from the previous activity and compare how the clusters were determined.

In Activity 07 I determined that the optimal number of clusters was 3 using the
Elbow method. However in this activity I found that they were actually 4 and 5
for the Single Linkage and Ward's Methods, respectively,

## 2. Are the clusters consistent?
If not, what do you think causes this, and how might it influence your
interpretation of biological patterns in the data?

The clusters are not consistent. I think that this has to do with the algorithms
we used. In Activity 07 we basically brute forced it and only checked up to k=10.
For this activity we instead used functions that are (probably) more robust and
statistically sound. The `hclust()` function must have been really well tested.

When it comes to interpreting biological patterns, even a difference of 1 in the
number of clusters (k) like 3 vs. 4 is significant. Clusters that are grouped
together when k=3 could be separate when k=4, for example. So things that you
consider "related" in K-means could be considered not related in Hierachical
Clustering. Another useful thing that HC provides is the ordering which you
could use to determine the hierarchy, hence its name. It reminds me of a
phlogenetic tree (like from Systematics) which is really useful to determine
how related any one point is from the others.

