---
title: 'Activity 13: Random Forest'
author: "Jerome Poblete"
date: "2026-07-23"
output: html_document
---

```{r setup, include=FALSE}
rm(list=ls())
knitr::opts_chunk$set(echo=TRUE)
```

# Base R Random Forest

## 2.) Read in the acoustic data
Remember to remove unnecessary columns (i.e. column 1).
Remember that your response variable is the species column.
```{r read-in-data}
call_data_df <- utils::read.csv("../data/call_data_12October2025.csv")
call_data_df <- call_data_df[complete.cases(call_data_df), -1]
```

## 3.) Design your cartesian grid.
Use the R `ranger` package to perform a random forest that iterates on each 
hyperparameter combination.

We can employ one (1) of two (2) methods.

### i. Classification Random Forest
- Used when the response variable is categorical (e.g. factor)
- Final prediction is based on **majority vote**
- Split criterion: *Gini impurity* (G) to measure node purity
    ```
    G = 1 - sum(p_i^2)
    
    where
    
    p_i = proportion of class i
    ```
- Smaller Gini = purer nodes

### ii. Regression Random Forest
- Used when the response variable is continuous
- Final prediction is based on **average** of all tree predictions
- Split criterion: Minimizes the *Residual Sum of Squares* (=RSS)
    ```
    RSS = sum((y_i - y_bar)^2)
    
    where
    y_i   = observed value
    y_bar = mean of node
    ```

```{r design-cartesian-grid}
library(dplyr)
library(ranger)

################################################################################
## A. Pick your poisons: predictor (x) and response (y) variables
################################################################################
x_df <- dplyr::select(call_data_df, Average_Frequency:Max_Amplitude)

## [1, ]: Species
y_df <- dplyr::select(call_data_df, Species)

## Number of observations
n <- nrow(call_data_df)

## Number of predictors
p <- ncol(x_df)


################################################################################
## B. Specify response variable
################################################################################
response_string <- "Species"

## Set hyperparamters
set.seed(123)
ntree <- 5

## Number of randomly selected predictors
mtry <- p |> sqrt() |> floor()


## Define the Gini impurity function
gini <- function(y) {
    ## Proportions of y
    p <- table(y_df) / length(y_df)
    return(1 - sum(p^2))
}


## Create a (zero-initialized) vector to store results, of size `ntree`.
forest_list <- vector("list", ntree)

## Zero-initialized numeric vector of size `ntree`.
oob_error_list <- numeric(ntree)

## Number literal 0 repeated `p` times.
importance_list        <- rep(0, p)
names(importance_list) <- names(x_df)
```

```{r gini-perform}
vprint <- function(..., sep = " ") print(paste(..., sep=sep))


## Build the format.
for (tree_index in 1:ntree) {
    vprint("Building Tree ", tree_index)
    
    ############################################################################
    ## C. Bootstrap the sample
    ############################################################################
    
    ## Take a sample from the input vector, of specified size, with replacement.
    boot_index   <- sample(1:n, size=n, replace=TRUE)
    boot_data_df <- call_data_df[boot_index, ]
    
    ## Out-of-bag observations
    ## `setdiff` is the difference between the two (2) sets.
    oob_index <- setdiff(1:n, unique(boot_index))
    
    ############################################################################
    ## D. Randomly select `mtry` predictors
    ############################################################################
    selected <- sample(names(x_df), mtry)
    
    vprint("Selected predictors:", paste(selected, collapse=", "))
    
    ############################################################################
    ## E. Evaluate all possible splits
    ############################################################################
    best_gini     <- Inf
    best_split    <- NA
    best_variable <- NA
    
    for (variable in selected) {
        ## One-dimensional (1-dimensional) vector of some predictor variable.
        predictor_df <- boot_data_df[[variable]]
        values       <- predictor_df |> unique() |> sort()
        if (length(values) < 2) {
            next
        }
        
        splits_list <- (values[-1] + values[-length(values)]) / 2
        for (split in splits_list) {
            ## Use the following to filter.
            left_df  <- boot_data_df[predictor_df <= split, ]
            right_df <- boot_data_df[predictor_df > split, ]
            
            ## No data points to work with?
            if (nrow(left_df) == 0 || nrow(right_df) == 0) {
                next
            }
            
            left_gini_df  <- gini(left$Species)
            right_gini_df <- gini(right$Species)
            weighted_gini <-
                (nrow(left_df)  / nrow(boot_data_df) * left_gini_df) +
                (nrow(right_df) / nrow(boot_data_df) * right_gini_df)
          
            ## Remember: lower Gini indicates purer nodes!
            if (weighted_gini < best_gini) {
                best_gini    <- weighted_gini
                best_split   <- split
                best_variable <- variable
            }
        }
    }

    ############################################################################
    ## F. Choose the best split
    ############################################################################
    vprint("Best variable :", best_variable)
    vprint("Best split    :", best_split)
    vprint("Weighted Gini :", best_gini)

    ############################################################################
    ## G. Grow the tree
    ##
    ## This example performs only the root split. A complete tree would
    ## recursively ## repeat the splitting process on each child node.
    ############################################################################
    ## Split the main data frame, row-wise, into left and right portions.
    best_variable_list  <- boot_data_df[[best_variable]]
    left_df             <- boot_data_df[best_variable_list <= best_split, ]
    right_df            <- boot_data_df[best_variable_list >  best_split, ]
    left_prediction     <- left_df$Species  |> table() |> which.max() |> names()
    right_prediction    <- right_df$Species |> table() |> which.max() |> names()
    
    vprint("Left node predicts  :", left_prediction)
    vprint("Right node predicts :", right_prediction)
    
    ## Save the tree.
    forest_list[[tree_index]] <- list(
        variable    = best_variable,
        split       = best_split,
        left_class  = left_prediction,
        right_class = right_prediction)

    ############################################################################
    ## J. Compute the Out-of-Bar error
    ############################################################################
    if (length(oob_index) > 0) {
        oob_df <- call_data_df[oob_index, ]
        prediction_list <- ifelse(
            oob_df[[best_variable]] <= best_split,
            left_prediction,
            right_prediction)
        
        ## `y` doesn't have a `levels` attribute, unlike the reference script.
        ## We don't need it anyway since the default behavior is the use the
        ## input as the basis for the levels, via `as.character() |> unique()
        ## |> sort()`.
        prediction_factor          <- factor(prediction_list)
        oob_error_list[tree_index] <- mean(prediction_factor != oob_df$Species)
        vprint("OOB Error ", tree_index, "=", oob_error_list[tree_index])
    }

    ############################################################################
    ## K. Variable importance
    ##
    ## Counts how many times each predictor becomes the best split.
    ############################################################################
    importance_list[best_variable] <- importance_list[best_variable] + 1
}

################################################################################
## H. Forest completed
################################################################################
vprint("Forest construction completed! Yeehaw!")
```

## 4.) Pick the best hyperparameter combination...
...that minimizes the out-of-bag error.
```{r combine-predictions}
################################################################################
## I. Combine predictions
################################################################################
shit       <- call_data_df[1, ]
votes_list <- character(ntree)

for (i in 1:ntree) {
    tree <- forest_list[[i]]
    if (shit[[tree$variable]] <= tree$split) {
        votes_list[i] <- tree$left_class
    } else {
        votes_list[i] <- tree$right_class
    }
}

final_prediction_string <- votes_list |> table() |> which.max() |> names()
vprint("Votes from each tree     :", paste(votes_list, collapse=", "))
vprint("Majority vote prediction :", final_prediction_string)
vprint("Average OOB error        :", mean(oob_error_list))
vprint("Variable importance      :", paste(importance_list, collapse=", "))
```


## 5.) Using the optimal hyperparameter...
...provide a plot of variable importance, and predicted class probabilities.
```{r plot-optimal-hyperparameters}
```