---
title: 'Activity 13: Random Forest'
author: "Jerome Poblete"
date: "2026-07-23"
output: html_document
---

```{r setup, include=FALSE}
rm(list=ls())
knitr::opts_chunk$set(echo=TRUE)
```

# Base R Random Forest

## 2.) Read in the acoustic data
Remember to remove unnecessary columns (i.e. column 1).
Remember that your response variable is the species column.
```{r read-in-data}
call_data_df <- utils::read.csv("../data/call_data_12October2025.csv")
call_data_df <- call_data_df[complete.cases(call_data_df), -1]
```

## 3.) Design your cartesian grid.
Use the R `ranger` package to perform a random forest that iterates on each 
hyperparameter combination.

We can employ one (1) of two (2) methods.

### i. Classification Random Forest
- Used when the response variable is categorical (e.g. factor)
- Final prediction is based on **majority vote**
- Split criterion: *Gini impurity* (G) to measure node purity
```
G = 1 - sum(p_i^2)

where

p_i = proportion of class i
```
- Smaller Gini = purer nodes

### ii. Regression Random Forest
- Used when the response variable is continuous
- Final prediction is based on **average** of all tree predictions
- Split criterion: Minimizes the *Residual Sum of Squares* (=RSS)
```
RSS = sum((y_i - y_bar)^2)

where
y_i   = observed value
y_bar = mean of node
```

```{r classification-random-forest}
library(dplyr)
library(ranger)

################################################################################
## A. Pick your poisons: predictor (x) and response (y) variables
################################################################################
## Predictor variables
x_df <- dplyr::select(call_data_df, Average_Frequency:Max_Amplitude)

## Number of observations
n <- nrow(call_data_df)

## Number of predictors
p <- ncol(x_df)


################################################################################
## B. Specify response variable
################################################################################
response_string <- "Species"

## `y` by default doesn't have a `levels` attribute, unlike the reference
## script, so convert it to a factor so it has one.
call_data_df[[response_string]] <- factor(call_data_df[[response_string]])

## [1, ]: Species
y_df  <- dplyr::select(call_data_df, dplyr::all_of(response_string))

## Necessary because we can't do do primitive comparisons, e.g. `==`, otherwise.
y_vec <- call_data_df[[response_string]]

response_formula <- Species ~ Average_Frequency +
    Average_Wavelength +
    Min_Amplitude +
    Max_Amplitude

################################################################################
## B.1. Set hyperparamters
################################################################################
set.seed(123)

## Number of trees
ntree <- 5

## Number of randomly selected predictors
mtry  <- p |> sqrt() |> floor()
```

## 4.) Pick the best hyperparameter combination...
...that minimizes the out-of-bag error.
```{r pick-best-hyperparameters}
## Build the random forest
rg_model <- ranger::ranger(response_formula,
    data       = call_data_df,
    num.trees  = ntree,
    mtry       = mtry,
    importance = "impurity")

## View model
rg_model

## Combine predictions (Majority vote)
## NOTE: Delegates to `predict.ranger()`.
rg_predictions <- stats::predict(rg_model, call_data_df)
head(rg_predictions)

## Confusion matrix
## Unlike with randomForest, we don't use the `predictions` object itself,
## because we want to compare with the underlying predicted responses.
table(Actual=y_vec, Prediction=rg_predictions$predictions)

## Classification Accuracy
rg_accuracy <- mean(rg_predictions$predictions == y_vec)
rg_accuracy

rg_model$prediction.error
```

## 5.) Optimal hyperparameter plots
### i.) Variable importance plot
```{r ranger-variable-importance-plot}
library(ggplot2)

## Variable importance
ranger::importance(rg_model)

## https://stackoverflow.com/questions/49105358/plot-feature-importance-computed-by-ranger-function
ggplot2::ggplot(
    data.frame(
        Variable   = names(rg_model$variable.importance),
        Importance = as.vector(rg_model$variable.importance)),
    aes(stats::reorder(Variable, Importance), Importance, fill = Importance)
) + ggplot2::geom_bar(stat = "identity", position = "dodge") + 
    ggplot2::coord_flip() +
    ggplot2::labs(
        title    = paste0("Variable Importance (", response_string, ")"),
        x        = "",
        y        = "Importance") +
    ggplot2::scale_fill_gradient(low = "red", high = "blue")
```

### ii.) Predicted class probability plot
```{r plot-optimal-hyperparameters}
library(rpart)
library(rpart.plot)

rg_tree <- rpart::rpart(response_formula, call_data_df)
rpart.plot::rpart.plot(rg_tree)
```

```{r design-cartesian-grid, echo=FALSE, eval=FALSE}
library(dplyr)

## Define the Gini impurity function
gini <- function(y_df) {
    ## Proportions of y
    p <- table(y_df) / length(y_df)
    return(1 - sum(p^2))
}


## Create a (zero-initialized) vector to store results, of size `ntree`.
forest_vec <- vector("list", ntree)

## Zero-initialized numeric vector of size `ntree`.
oob_error_vec <- numeric(ntree)

## Number literal 0 repeated `p` times.
importance_vec        <- rep(0, p)
names(importance_vec) <- names(x_df)
```

```{r gini-perform, echo=FALSE, eval=FALSE}
vprint <- function(..., sep = " ") print(paste(..., sep=sep))

################################################################################
## C. Bootstrap the sample
################################################################################
bootstrap_sample <- function(data_df, n) {
    ## Take a sample from the input vector, of specified size, with replacement.
    boot_index   <- sample(1:n, size=n, replace=TRUE)
    boot_data_df <- data_df[boot_index, ]
    
    ## Out-of-bag observations
    ## `setdiff` is the difference between the two (2) sets.
    oob_index <- setdiff(1:n, unique(boot_index))
    return(list(data_df=boot_data_df, oob_index=oob_index))
}


evaluate_splits <- function(boot, x_df, mtry) {
    ############################################################################
    ## D. Randomly select `mtry` predictors
    ############################################################################
    selected <- sample(names(x_df), mtry)
    # vprint("Selected predictors:", paste(selected, collapse=", "))
    
    
    ############################################################################
    ## E. Evaluate all possible splits
    ############################################################################
    best_gini     <- Inf
    best_split    <- NA
    best_variable <- NA
    for (variable in selected) {
        ## One-dimensional (1-dimensional) vector of some predictor variable.
        predictor_df <- boot$data_df[[variable]]
        values       <- predictor_df |> unique() |> sort()
        if (length(values) < 2) {
            next
        }
        
        splits_vec <- (values[-1] + values[-length(values)]) / 2
        for (split in splits_vec) {
            ## Use the following to filter.
            left_df  <- boot$data_df[predictor_df <= split, ]
            right_df <- boot$data_df[predictor_df > split, ]
            
            ## No data points to work with?
            if (nrow(left_df) == 0 || nrow(right_df) == 0) {
                next
            }
            
            left_gini_df  <- gini(left_df[[response_string]])
            right_gini_df <- gini(right_df[[response_string]])
            weighted_gini <-
                (nrow(left_df)  / nrow(boot$data_df) * left_gini_df) +
                (nrow(right_df) / nrow(boot$data_df) * right_gini_df)
          
            ## Remember: lower Gini indicates purer nodes!
            if (weighted_gini < best_gini) {
                best_gini     <- weighted_gini
                best_split    <- split
                best_variable <- variable
            }
        }
    }
    
    
    ############################################################################
    ## F. Choose the best split
    ############################################################################
    return(list(
        variable = best_variable,
        data_vec = boot$data_df[[best_variable]],
        gini     = best_gini,
        split    = best_split))
}


############################################################################
## G. Grow the tree
##
## This example performs only the root split. A complete tree would
## recursively ## repeat the splitting process on each child node.
##
## NOTE: We can't passs `forest_vec` by reference, so we have to mutate
## it explicitly.
############################################################################
grow_tree <- function(boot, best, tree_index) {
    ## Split the main data frame, row-wise, into left and right portions.
    left_df             <- boot$data_df[best$data_vec <= best$split, ]
    right_df            <- boot$data_df[best$data_vec >  best$split, ]

    ## doozy
    left_prediction <- left_df[[response_string]] |>
        table()     |>
        which.max() |>
        names()
    right_prediction <- right_df[[response_string]] |>
        table()     |>
        which.max() |>
        names()

    # vprint("Left node predicts  :", left_prediction)
    # vprint("Right node predicts :", right_prediction)
    
    tree <- list(
        variable    = best$variable,
        split       = best$split,
        left_class  = left_prediction,
        right_class = right_prediction)

    ## Save the tree. `<<-` assigns in the parent scope.
    forest_vec[[tree_index]] <<- tree
    return(tree)
}


################################################################################
## J. Compute the Out-of-Bar error
################################################################################
compute_oob_error <- function(boot, best, tree_index, tree) {
    if (length(boot$oob_index) > 0) {
        oob_df <- call_data_df[boot$oob_index, ]
        prediction_factor <- (ifelse(
            oob_df[[best$variable]] <= best$split,
            tree$left_class,
            tree$right_class))
        
        ## Save this in the parent scope using the `<<-` operator.
        oob_error_vec[tree_index] <<- mean(prediction_factor != oob_df$Species)
        # vprint("OOB Error ", tree_index, "=", oob_error_vec[tree_index])
    }
}


## Build the format.
for (tree_index in 1:ntree) {
    boot <- bootstrap_sample(call_data_df, n)
    best <- evaluate_splits(boot, x_df, mtry)
    tree <- grow_tree(boot, best, tree_index)
    compute_oob_error(boot, best, tree_index, tree)

    ############################################################################
    ## K. Variable importance
    ##
    ## Counts how many times each predictor becomes the best split.
    ############################################################################
    importance_vec[best$variable] <- importance_vec[best$variable] + 1
}

################################################################################
## H. Forest completed
################################################################################
```

```{r combine-predictions, echo=FALSE, eval=FALSE}
################################################################################
## I. Combine predictions
################################################################################
combine_predictions <- function(data_df, ntree) {
    new_data_df <- data_df[1, ]
    votes_vec   <- character(ntree)
    for (i in 1:ntree) {
        tree         <- forest_vec[[i]]
        votes_vec[i] <- ifelse(
            new_data_df[[tree$variable]] <= tree$split,
            tree$left_class,
            tree$right_class)
    }
    return(votes_vec)
}

votes_vec        <- combine_predictions(call_data_df, ntree)
final_prediction <- votes_vec |> table() |> which.max() |> names()
vprint("Votes from each tree     :", paste(votes_vec, collapse=", "))
vprint("Majority vote prediction :", final_prediction)
vprint("Average OOB error        :", mean(oob_error_vec))
vprint("Variable importance      :", paste(importance_vec, collapse=", "))
```

```{r predict-randomForest, echo=FALSE, eval=FALSE}
library(randomForest)

rf_model <- randomForest::randomForest(response_formula,
    data       = call_data_df,
    ntree      = ntree,
    mtry       = mtry,
    importance = TRUE)

##View model
vprint(rf_model)
```

```{r randomForest-get-optimal-hyperparameters, echo=FALSE, eval=FALSE}
## Temp.
rf_model <- randomForest::randomForest(response_formula,
    data       = call_data_df,
    ntree      = ntree,
    mtry       = mtry,
    importance = TRUE)

## View model
rf_model

## Combine predictions (Majoirty vote)
## Delegates to `predict.randomForest()`.
rf_predictions <- stats::predict(rf_model, call_data_df)
head(rf_predictions)

## Confusion matrix
table(Actual=y_vec, Prediction=rf_predictions)

## Classification accuracy
accuracy_rf <- mean(rf_predictions == y_vec)
accuracy_rf

rf_model$err.rate[ntree, "OOB"]
```

```{r randomForest-plot-variable-importance, echo=FALSE, eval=FALSE}
## Variable importance
randomForest::importance(rf_model)

## Plot variable importance
randomForest::varImpPlot(rf_model)

## Plot OOB error versus number of trees
plot(rf_model$err.rate[, "OOB"],
    type = "l",
    col  = "black",
    lwd  = 2,
    ylim = c(0, max(rf_model$err.rate)),
    xlab = "Number of Trees",
    ylab = "Error Rate",
    main = "Out-of-Bag Error vs. Number of Trees")

## Add class-specific OOB error lines
rf_names <- colnames(rf_model$err.rate)
graphics::lines(rf_model$err.rate[, rf_names[[2]]], col="red",   lwd=2)
graphics::lines(rf_model$err.rate[, rf_names[[3]]], col="green", lwd=2)
graphics::lines(rf_model$err.rate[, rf_names[[4]]], col="blue",  lwd=2)

## Add a legend
graphics::legend("topright",
    legend = rf_names,
    col    = c("black", "red", "green", "blue"),
    lwd    = 2,
    bty    = "n")

```
