---
title:  "Activity 09: Dimension Reduction: Principal Components Analysis (PCA)"
author: "Jerome Poblete"
date:   "2026 June 18"
output: html_document
---

Setup for later!
```{r setup-colors, echo=FALSE}
# https://github.com/joshdick/onedark.vim/blob/main/colors/onedark.vim
atom.onedark <- data.frame(
  black    = "#282C34",
  lightred = "#E06C75",
  green    = "#98C379",
  yellow   = "#E5C07B",
  blue     = "#61AFEF",
  magenta  = "#C678DD",
  cyan     = "#56B6C2",
  white    = "#ABB2BF",
  darkred  = "#BE5046"
)
```

# A. Principal Components
## 2. Read in the frogs' data as a dataframe
```{r read-in}
frog.data <- read.csv("../data/Meneses_et.al_2026_Platymantis.csv")
```

## 3. Subset your data to retain only the "Species" column and morphological features
E.g. SVL, HL, SNL, etc., from columns 10 through the last one (i.e. 10:X).

```{r subset-data}
frog.numeric <- frog.data[, 10:ncol(frog.data)]
```

## 4. Check the class & structure of the subset data. How many morphological features are present in the dataset?
```{r check-class}
library(tidyverse)

class(frog.numeric)
structure(frog.numeric)
frog.morph <- frog.numeric |> dplyr::select(10:ncol(frog.numeric))
length(frog.morph)
```

There are 16 morphological feature present in the data set.

## 5. Ensure the data frame has no NAs.
```{r ensure-data}
frog.numeric <- frog.numeric[frog.numeric |> complete.cases(), ]
frog.morph   <- frog.morph[frog.morph |> complete.cases(), ]
```

## 6. Scale the data using only the numeric columns, then compute the pairwise correlations. Do not further subset the data.
```{r scale-data}
frog.scaled <- frog.morph  |> scale()
frog.cmat   <- frog.scaled |> stats::cor() |> round(2)
```

## 7. Use the correlation matrix to...
### i. Calculate the following:
#### a. Eigenvectors
```{r eigenvectors}
frog.eig <- frog.cmat |> eigen()
frog.eig
```

#### b. Proportion of variance (percentage)
```{r proportion-of-variance}
pvar <- function(x) {
  props <- x / sum(x)
  data.frame(Proportion = props, Percentage = props * 100)
}

frog.pvar <- pvar(frog.eig$values)
frog.pvar |> round(4)
```

### ii. How many principal components were produced?
```{r count-principal-components}
## PC1, PC2, ... PCi
pc_strings <- paste0("PC", 1:length(frog.pvar$Proportion))

## %*% is the matrix multiplication operator.
frog.scores <- data.frame(frog.scaled %*% frog.eig$vectors)
colnames(frog.scores) <- pc_strings

frog.scores.head <- frog.scores |> head()
frog.scores.head |> round(4)
frog.scores.head |> nrow()
```

6 principals components were produced.

### iii. What is the proportion of variance explained by components 1-4?
```{r}
frog.pvar$Percentage[1:4] |> round(2)
```

Components 1 through 4 explain 33.91%, 16.09%, 11.18%, and 9.34% of the
variance, respectively.

## 8. Create a 1x2 plot showing screeplot & the PCA biplot
You can use either base R or ggplot2.
You can be creative here by coloring the point shapes per species and adding a
legend.
Consider the high data-ink ratio you learned from the lecture.

### i. Scree Plot
#### a. Base R
```{r scree-plot-base, echo=FALSE}
graphics::barplot(frog.pvar$Proportion,
  main      = "Frog Morphology: Explained Variance (Scree Plot)",
  xlab      = "Principal Component", ylab = "Proportion of Variance",
  sub       = "Base R version",
  names.arg = pc_strings,  # x-axis labels should be of the correct count.
  ylim      = c(0.0, 1.0), # y-axis values are in the range [0.0, 1.0].
  border    = NA,
  col       = atom.onedark$cyan)
```

#### b. ggplot2
```{r scree-plot-ggplot2, echo=FALSE}
library(ggplot2)

# `factor()` with the 2nd argument `levels` is needed to ensure the
# original ordering we passed in. Otherwise ggplot will arrange it by
# ASCII value.
ggplot2::ggplot(frog.pvar,aes(pc_strings |> factor(pc_strings), Proportion)) +
  # Flipped version of `ggplot2::geom_bar()`.
  ggplot2::geom_col(fill = atom.onedark$magenta) +
  # Add these so the plot is a little more independent.
  ggplot2::labs(x="Principal Component", y="Proportion of Variance",
    title    = "Frog Morphology: Explained Variance (Scree Plot)",
    subtitle = "ggplot2 Version") +
  # Clamp %variance.
  ggplot2::lims(y = c(0.0, 1.0))
```

### ii. PCA biplot
#### a. Base R
```{r pca-biplot-base, echo=FALSE}
# Map indexes to species as a `factor`, as character vectors are not useful
# for plotting purposes. This helps us determine which score is of what species.
frog.species  <- frog.data$Species |> factor()

mycolors <- c(atom.onedark$lightred, atom.onedark$blue)

# Must be set BEFORE the plot is drawn!
# We need this because `plot(bg=)` is NOT the same as this.
# par(fg       = atom.onedark$white,
#     col.axis = atom.onedark$white,
#     col.lab  = atom.onedark$white,
#     col.main = atom.onedark$white,
#     col.sub  = atom.onedark$white,
#     bg       = atom.onedark$black)

graphics::plot(frog.scores[,1], frog.scores[,2],
  main = "Frog Morphology: PCA Score Plot (PC1 vs. PC2)",
  xlab = "PC1", ylab = "PC2",
  pch  = 19,
  # We need a vector of colors, not an array of factors.
  # NOTE The `labels` argument of `factor` describes *what* to use based
  #      on treating each index as a level.
  col  = frog.data$Species |> factor(labels=mycolors) |> as.character())

graphics::legend("bottomleft",
  title  = "Species",
  legend = frog.species |> levels(),
  col    = mycolors, # Color of points/lines in the legend.
  pch    = 19)

# Add eigenvector arrows.
load_scale <- 7.5 # Adjust for visibility
loadings_scaled <- frog.eig$vectors[, 1:2] * load_scale
graphics::arrows(0, 0,
  loadings_scaled[, 1], loadings_scaled[, 2],
  length = 0.10, # Arrow head size.
  col =  atom.onedark$black,
  lty = "solid", # Line type. See ?par.
  lwd = 1.25)       # Line width. Must be >= 1.

## Add labels for the variables.
graphics::text(loadings_scaled[, 1], loadings_scaled[, 2],
  labels = colnames(frog.morph),
  col    = atom.onedark$black,
  pos    = 3, # 1 = below, 2= left, 3 = above, 4 = right
  cex    = 0.75) # Numeric character expansion factor.
```

#### b. ggplot2
```{r pca-biplot-gglot2, echo = FALSE}
ggplot2::ggplot(frog.scores, aes(PC1, PC2, colour=frog.species)) +
  ggplot2::geom_point() +
  # This is how we can customize the colors of the grouping variable.
  # Each color maps to a particular level. This is NOT a map from individual
  # data points to colors.
  ggplot2::scale_colour_manual(values = mycolors) +
  ggplot2::labs(title= "Frog Morphology: PCA Score Plot (PC1 vs. PC2)",
       subtitle = "ggplot2 Version",
       # Affects the legend title, confusingly enough!
       # See: https://r-charts.com/ggplot2/legend/
       colour   = "Species") +
  ggplot2::theme(legend.position = "right")
```

# B. Broken-stick vs. PCAtest
## 1. Determine the variance explained by principal components
### i. Broken-stick method
```{r broken-stick}
## p = #variables
## k = component index
broken_stick <- function(p) {
  sapply(1:p, function(k) {sum(1 / (k:p)) / p})
}

frog.bs <- broken_stick(ncol(frog.morph))
frog.bs |> round(4)

## Combine observed and broken stick variance.
frog.cmp <- data.frame(
  Observed     = frog.pvar$Proportion,
  Broken_Stick = frog.bs,
  row.names    = pc_strings
)
frog.cmp |> round(4)
```

#### a. Base R plot
```{r broken-stick-base, echo = FALSE}
graphics::plot(frog.pvar$Proportion,
  main  = "Frog Morphology: PCA (Broken-Stick Analysis)",
  xlab  = "Principal Component", ylab = "Proportion of Variance",
  type  = "b",
  pch   = 19,
  col   = atom.onedark$cyan,
  ylim  = c(0, max(frog.pvar$Proportion)))

graphics::lines(frog.bs,
  type = "b",
  pch  = 17,
  col  = atom.onedark$lightred)

graphics::legend("topright",
  legend = c("Observed", "Broken Stick"),
  pch    = c(19, 17),
  col    = c(atom.onedark$cyan, atom.onedark$lightred))
```

### ii. PCAtest package

```{r pca-test}
# install.packages("devtools")
# devtools::install_github("arleyc/PCAtest")
library(PCAtest)

frog.pca <- frog.morph |> PCAtest::PCAtest(100, 100, 0.05, counter=FALSE, plot=TRUE)
```

### iii. Q: What is the difference between these 2 methods for assessing the importance of principal components?

The variances explained by the 1st 4 principal components of the Broken-Stick
method are 21.13%, 14.88%, 11.75%, and 9.67%. Most of these are lower than
their respective PCs in the Observed data, except for PC3. The Broken-Stick
method we used also notably uses only the number of principal components
to calculate the PCs. Everything is effectively just in relation to the
numeroc of PCs we started with.

The PCAtest library, according to its help (see `?PCAtest`), uses random
permutations of the input data. It also helps us determines exactly which
PCs are indeed significant. This probably means it is more reliable, although
more computationally intensive.


## 2. Create a barplot showing the results of Broken-stick analysis
Also perform a significance test for principal components using PCAtest,
following the instructions at https://arleyc.github.io/PCAtest/.

From the previous item we know that only the 1st 4 PCs are significant.

### i. Base R

```{r broken-stick-graphics-barplot, echo=FALSE}

# Treat significant PCs as green and the rest as red.
frog.bs_colors <- rep(atom.onedark$lightred, length(frog.bs))
frog.bs_colors[1:4] <- atom.onedark$green

graphics::barplot(frog.bs,
  main      = "Frog Morphology: PCA (Broken-Stick)",
  sub       = "Base R version",
  xlab      = "Principal Component", ylab = "Proportion of Variance",
  names.arg = pc_strings,
  col       = frog.bs_colors,
  border    = NA,
  ylim      = c(0.0, 1.0))
```


### ggplot2

```{r broken-stick-ggplot-barplot, echo=FALSE}
ggplot2::ggplot(frog.cmp, aes(pc_strings |> factor(pc_strings), Broken_Stick)) +
  ggplot2::geom_col(fill = frog.bs_colors) +
  ggplot2::labs(x="Principal Component", y="Proportion of Variance",
    title    = "Frog Morphology: PCA (Broken-Stick)",
    subtitle = "ggplot2 Version") +
  ggplot2::lims(y = c(0.0, 1.0))
```