---
title: 'Activity 10: Geographic Information System'
author: "Jerome Poblete"
date: "2026-06-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# A. Retrieving spatial data
## 1. In RStudio, create an R Markdown file. Name it “Your Surname_gis.rmd.”

You're looking at it right now!

## 2. Go to the Global Biodiversity Information Facility (GBIF) website.
Create your account and download your assigned species records below as a CSV file (see Table).

## 3. Clean the data by removing records found outside the reported species range.

```{r clean-data}
library(tidyverse)

# TODO: Do we want to include collection infromation? e.g. date, collectors?
gbif <- read.delim("../data/Barbourula_busuangensis_GBIF.csv")
gbif.coords <- gbif |> dplyr::select(decimalLatitude:decimalLongitude)
gbif.coords <- gbif.coords[gbif.coords |> complete.cases(), ]
```

## 4. Go to the DIVA-GIS website and download the Philippine administrative boundaries/areas.
The zip file contains multiple shapefiles (adm0–3), representing different boundary levels (0-country; 1-provincial; 2-municipal; 3-barangay).

```{r diva-gis-download}
library(sf) # For reading shapefiles.
phl_adm <- sf::st_read("../data/PHL_adm/PHL_adm1.shp")
```

## 5. Get only the provincial level, and subset the provinces that contain only distribution records for your assigned species.

```{r diva-gis-subset}
gbif.provinces <- gbif |> dplyr::select(stateProvince) |> unique()
phl_adm.palawan <- phl_adm[59, ] # Palawan is row 59.
```

## 6. Go back to the DIVA-GIS website.
Download the Philippine digital elevation model (“Elevation [country mask]”), then clip it using the provinces you selected in step 5.

```{r diva-gis-elevation}
```

## 7. If DIVA-GIS is acting weirdly, you can download the admin boundaries here and elevation here.

# B. Create your first map
## 1. Now, put all the data together into a map, following the pre-lab lecture.
Make sure that data layers (points, polygons, DEMs) are arranged properly. Be creative!

