---
title: "Poblete_Lasquite_gradient_boost"
author: "Hannah Lasquite and Jerome Poblete"
output: html_document
---

# 1. What is Gradient Forest?
# 2. Gradient Boosting vs. Random Forest
# 3. Implementation of Gradient Boosting in Different Data Types
# 4. Regression Gradient Boosting Without Packages

# Step 1: Cleaning up our data, the in-built penguins dataset, and setting aside training and testing data.

```{r S41, warning = FALSE}
#Including only complete rows of data
clean_penguin <- na.omit(penguins)

#Randomize all the rows and split them for testing and training

set.seed(123)
random_penguin <- clean_penguin[sample(nrow(clean_penguin)), ]

train_penguin <- random_penguin[1:233, ]  #70% of the dataset
test_penguin <- random_penguin[234:333, ] #30% of the dataset
```

# Step 2: Assigning variables
```{r S42, warning = FALSE}
X <- train_penguin[, c("bill_len", "bill_dep", "flipper_len")]
Y <- train_penguin$body_mass

obN <- nrow(train_penguin)

df <- data.frame(X,Y)

pred1 <- mean(Y)
pred2 <- rep(pred1, obN)
```

# Step 3: Gradient Boosting Loop
```{r S43, warning = FALSE}
n_trees <- 50
learning_rate <- 0.1
model <- list()

for (i in 1:n_trees) {
    residuals <- df$Y - pred2 #Residual for each row
    inttree <- lm(residuals ~ bill_len + bill_dep + flipper_len, data = df)
    
    update_pred <- predict(inttree, newdata = df)
  pred2 <- pred2 + (learning_rate * update_pred) #Update the predictions while its being scaled by the learning rate
   model[[i]] <- inttree
}

predictdata <- function(new_data, model, pred1, learning_rate){
  pred3 <- rep(pred1, nrow(new_data))
  for (m in model) {
    pred3 <- pred3 + (learning_rate * predict(m, new_data = newdata))
  }
  return(pred3)
}

final_preds <- predictdata(df, model, pred1, learning_rate)
cat("Predicted Body Mass:\n", head(final_preds),"\n", "\n")
cat("Actual Body Mass:\n", head(Y))
```
# Step 4: Testing on New Dataset
```{r S44, warning = FALSE}
X2 <- test_penguin[, c("bill_len", "bill_dep", "flipper_len")]
Y2 <- test_penguin$body_mass

obN2 <- nrow(test_penguin)

df2 <- data.frame(X2,Y2)

model2 <- list()

pred_1 <- mean(Y2)

final_preds2 <- predictdata(df2, model, pred_1, learning_rate)
cat("Predicted Body Mass:\n", head(final_preds2),"\n", "\n")
cat("Actual Body Mass:\n", head(Y2))
```
# Conclusion:
*A penguin's bill depth, bill length, or flipper length are not accurate markers for it's overall body mass across different species*